---
title: "Meta Data"
execute: 
  warning: false
  message: false
fig-pos: H
format:
  pdf:
    colorlinks: true
    pdf-engine: xelatex
---

# Data

## Fern General

### Functions

```{r}
library(readr)
library(dplyr)
library(here)

aggregate_csvs <- function(path_parts,
                           output_name = "aggregated.csv",
                           dedupe = TRUE) {
  
  # Build directory path using here()
  data_dir <- do.call(here::here, as.list(path_parts))
  
  cat("Reading from:", data_dir, "\n")
  
  if (!dir.exists(data_dir)) {
    stop("Directory does not exist: ", data_dir)
  }
  
  # Get CSV files
  csv_files <- list.files(
    path = data_dir,
    pattern = "\\.csv$",
    full.names = TRUE
  )
  
  if (length(csv_files) == 0) {
    stop("No CSV files found in: ", data_dir)
  }
  
  cat("Found", length(csv_files), "files\n")
  
  # Read and combine
  combined_data <- lapply(csv_files, function(f) {
    read_csv(f, show_col_types = FALSE) %>%
      mutate(source_file = basename(f))
  }) %>%
    bind_rows()
  
  # Deduplicate (optional)
  if (dedupe) {
    combined_data_dedup <- combined_data %>% distinct()
  } else {
    combined_data_dedup <- combined_data
  }
  
  # Output path at project root
  output_dir <- here::here("outputs")
  
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  output_file <- file.path(output_dir, output_name)
  
  # Write file
  write_csv(combined_data_dedup, output_file)
  
  # Logs
  cat("Saved to:", output_file, "\n")
  cat("Rows before dedupe:", nrow(combined_data), "\n")
  cat("Rows after dedupe:", nrow(combined_data_dedup), "\n")
  
  return(combined_data_dedup)
}
```

```{r}
analyze_dataset <- function(df) {
  list(
    missing = df %>%
      summarise(across(everything(), ~mean(is.na(.)))),

    numeric_summary = df %>%
      summarise(across(where(is.numeric),
                       list(mean = mean, sd = sd),
                       na.rm = TRUE)),

    n_rows = nrow(df),
    n_cols = ncol(df)
  )
}
```

### Business Activity

```{r}
ba_data <- aggregate_csvs(
  path_parts = c("Fern_General", "Business_Activity"),
  output_name = "business_activity_aggregated.csv"
)
```

### Movement Between Places

```{r}
mp_data <- aggregate_csvs(
  path_parts = c("Fern_General", "Movement_between_Places"),
  output_name = "movement_between_places_aggregated.csv"
)
```

### Summary Statistics

```{r}
analyze_dataset(ba_data)
analyze_dataset(mp_data)
```

## Tennessee_Kentucky

### Movement Between Places During a Crisis

```{r}
tk_mp_data_admin <- aggregate_csvs(
  path_parts = c("Tennessee_Kentucky", "Fern_Movement_Crisis","Administrative_Regions"),
  output_name = "tk_movement_between_places_aggregated_admin.csv"
)
```

```{r}
tk_mp_data_bing <- aggregate_csvs(
  path_parts = c("Tennessee_Kentucky", "Fern_Movement_Crisis","Bing_Tiles"),
  output_name = "tk_movement_between_places_aggregated_bing.csv"
)
```

### Facebook Population during crisis

```{r}
tk_fb_data_admin <- aggregate_csvs(
  path_parts = c("Tennessee_Kentucky", "Fern_Facebook_Pop_Crisis","Administrative_Regions"),
  output_name = "tk_facebook_pop_aggregated_admin.csv"
)
```

```{r}
tk_fb_data_bing <- aggregate_csvs(
  path_parts = c("Tennessee_Kentucky", "Fern_Facebook_Pop_Crisis","Bing_Tiles"),
  output_name = "tk_facebook_pop_aggregated_bing.csv"
)
```

### Network Coverage

```{r}
tk_nc_data_active <- aggregate_csvs(
  path_parts = c("Tennessee_Kentucky", "Fern_Network_Coverage","Active"),
  output_name = "tk_network_coverage_aggregated_active.csv"
)
```

```{r}
tk_nc_data_undetected <- aggregate_csvs(
  path_parts = c("Tennessee_Kentucky", "Fern_Network_Coverage","Undetected"),
  output_name = "tk_network_coverage_aggregated_undetected.csv"
)
```

## Latitude and Longitude

```{r}
# convert latitude/longitude points into sf objects, then join them to county polygons
library(sf)
library(dplyr)
library(tigris)

options(tigris_use_cache = TRUE)

# Get US county boundaries
counties_sf <- counties(cb = TRUE, year = 2023) %>%
  st_transform(4326) %>%
  select(
    county_state = STATE_NAME,
    county_name = NAME,
    county_geoid = GEOID,
    geometry
  )
```

```{r}
# Assign coordinates to counties
add_county_from_coords <- function(df, lat_col, lon_col, prefix = "") {
  
  points_sf <- df %>%
    filter(!is.na(.data[[lat_col]]), !is.na(.data[[lon_col]])) %>%
    st_as_sf(
      coords = c(lon_col, lat_col),
      crs = 4326,
      remove = FALSE
    )
  
  joined <- st_join(points_sf, counties_sf, join = st_within)
  
  joined %>%
    st_drop_geometry() %>%
    rename(
      "{prefix}county_state" := county_state,
      "{prefix}county_name" := county_name,
      "{prefix}county_geoid" := county_geoid
    )
}
```

## General

### Business Activity

```{r}
ba_data <- add_county_from_coords(
  df = ba_data,
  lat_col = "latitude",
  lon_col = "longitude",
  prefix = ""
)
```

### Movement Between Places (In progress/runtime issues)

```{r, execute = FALSE}
# Origin
mp_data<- add_county_from_coords(
  df = mp_data,
  lat_col = "start_latitude",
  lon_col = "start_longitude",
  prefix = "start_"
)

# Destination
mp_data_counties <- add_county_from_coords(
  df = mp_data_origin_counties,
  lat_col = "end_latitude",
  lon_col = "end_longitude",
  prefix = "end_"
)

```

## Tennessee Kentucky

### Movement Between Places

```{r}
# Administrative Regions
# Origin counties
tk_mp_data_admin <- add_county_from_coords(
  df = tk_mp_data_admin,
  lat_col = "start_latitude",
  lon_col = "start_longitude",
  prefix = "start_"
)

# Destination counties
tk_mp_data_admin <- add_county_from_coords(
  df = tk_mp_data_admin,
  lat_col = "end_latitude",
  lon_col = "end_longitude",
  prefix = "end_"
)

# Bing Tiles
# Origin counties
tk_mp_data_bing <- add_county_from_coords(
  df = tk_mp_data_bing,
  lat_col = "start_latitude",
  lon_col = "start_longitude",
  prefix = "start_"
)

# Destination counties
tk_mp_data_bing <- add_county_from_coords(
  df = tk_mp_data_bing,
  lat_col = "end_latitude",
  lon_col = "end_longitude",
  prefix = "end_"
)
```

### Facebook Population

```{r}
# Administrative Regions

tk_fb_data_admin <- add_county_from_coords(
  df = tk_fb_data_admin,
  lat_col = "latitude",
  lon_col = "longitude",
  prefix = ""
)

# Bing Tiles

tk_fb_data_bing <- add_county_from_coords(
  df = tk_fb_data_bing,
  lat_col = "latitude",
  lon_col = "longitude",
  prefix = ""
)


```

### Network Coverage

```{r}
# Active
tk_nc_data_active <- add_county_from_coords(
  df = tk_nc_data_active,
  lat_col = "latitude",
  lon_col = "longitude",
  prefix = ""
)

tk_nc_data_undetected <- add_county_from_coords(
  df = tk_nc_data_undetected,
  lat_col = "latitude",
  lon_col = "longitude",
  prefix = ""
)
```

# Visualizations

Based on LA Fire [Situation Report](https://www.crisisready.io/wp-content/uploads/2025/01/ReadyMapper-LAFires-01152025.pdf)

```{r}
library(ggplot2)
library(gt)
library(scales)
library(lubridate)

latest_date <- max(tk_fb_data_admin$ds, na.rm = TRUE)

fb_latest_county <- tk_fb_data_admin %>%
  filter(ds == latest_date) %>%
  group_by(county_geoid, county_name, county_state) %>%
  summarise(
    mean_percent_change = mean(percent_change, na.rm = TRUE),
    baseline_pop = sum(n_baseline, na.rm = TRUE),
    crisis_pop = sum(n_crisis, na.rm = TRUE),
    .groups = "drop"
  )

fb_map <- counties_sf %>%
  inner_join(fb_latest_county, by = "county_geoid")

ggplot(fb_map) +
  geom_sf(aes(fill = mean_percent_change), linewidth = 0.2) +
  scale_fill_gradient2(
    labels = label_number(suffix = "%", accuracy = 1)
  ) +
  labs(
    title = paste("Population Density Change on", latest_date),
    fill = "% Change vs Baseline"
  ) +
  theme_minimal()
```
Check how we're getting 


