---
title: "Winter Storm Fern Situation Report - January 2026"
execute: 
  warning: false
  message: false
  output: false
  echo: false
fig-pos: H
format:
  pdf:
    colorlinks: true
    pdf-engine: xelatex
---


Formatted based on LA Fire [Situation Report](https://www.crisisready.io/wp-content/uploads/2025/01/ReadyMapper-LAFires-01152025.pdf)

# Summary

A major American winter storm, often referred to as Winter Storm Fern, started on Friday, January 23 and continued through January 26th. The storm brought heavy snow and freezing rain to several U.S. states, ranging from the southern plains to the Eastern United States. Several states experienced network outages and extremely cold temperatures.

The following visualizations are developed based on Meta AI for Good's data through Facebook. The data covers network coverage and population density from January 27 to February 10, with a focus on Tennessee and Kentucky and daily updates during the collection period.


# Movement

```{r}
library(ggplot2)
library(gt)
library(scales)
library(lubridate)
library(ggmap)

make_percent_change_map <- function(file_name,
                                    map_date = NULL,
                                    date_col = "ds",
                                    value_col = "percent_change",
                                    output_plot_name = NULL,
                                    title_prefix = "Population Density Change on",
                                    maptype = "stamen_toner_lite",
                                    zoom = 8,
                                    bbox_buffer = 0.25) {
  
  # Read file from root outputs folder
  df <- read_csv(
    here("Outputs", file_name),
    show_col_types = FALSE
  )
  
  # Make sure the date column is treated as a Date
  df <- df |>
    mutate("{date_col}" := as.Date(.data[[date_col]]))
  
  # If no date is supplied, use the latest date by default
  if (is.null(map_date)) {
    map_date <- max(df[[date_col]], na.rm = TRUE)
  } else {
    map_date <- as.Date(map_date)
  }
  
  # Summarize to county level for the selected date
  selected_county <- df |>
    filter(.data[[date_col]] == map_date) |>
    group_by(county_geoid, county_name, county_state) |>
    summarise(
      mean_percent_change = mean(.data[[value_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  if (nrow(selected_county) == 0) {
    stop("No data found for selected date: ", map_date)
  }
  
  # Join county summaries to county polygons
  map_data <- counties_sf |>
    inner_join(selected_county, by = "county_geoid") |>
    st_transform(4326)
  
  # Build bounding box around relevant counties
  bbox <- st_bbox(map_data)
  
  bbox_stadia <- c(
    left   = bbox["xmin"] - bbox_buffer,
    bottom = bbox["ymin"] - bbox_buffer,
    right  = bbox["xmax"] + bbox_buffer,
    top    = bbox["ymax"] + bbox_buffer
  )
  
  # Download Stadia basemap
  # get_stadiamap() expects bbox as lon/lat: left, bottom, right, top
  base_map <- get_stadiamap(
    bbox = bbox_stadia,
    zoom = zoom,
    maptype = maptype
  )
  
  # Plot county polygons over the basemap
  p <- ggmap(base_map) +
    geom_sf(
      data = map_data,
      aes(fill = mean_percent_change),
      inherit.aes = FALSE,
      color = "white",
      linewidth = 0.25,
      alpha = 0.8
    ) +
    scale_fill_gradient2(
      labels = label_number(suffix = "%", accuracy = 1)
    ) +
    labs(
      title = paste(title_prefix, map_date),
      fill = "% Change vs Baseline"
    ) +
    theme_minimal()
  
  # Save plot if requested
  if (!is.null(output_plot_name)) {
    ggsave(
      filename = here("Outputs", output_plot_name),
      plot = p,
      width = 10,
      height = 7
    )
  }
  
  return(p)
}
```


```{r}
# Not working because of earlier runtime error do not delete
make_percent_change_map(
  file_name = "movement_between_places_aggregated.csv"
)
```

## % Population Density Change

Note the difference in @fig-pop-bing, which is based on Bing tiles, and @fig-pop-admin, which is based on administrative regions.

```{r fig-pop-bing, fig.cap = "Based on the Bing tiles, there are only a few counties with strong population density changes.", output = T}
make_percent_change_map(
  file_name = "tk_facebook_pop_aggregated_bing.csv"
)
```



```{r, eval=F}
# Not working because of earlier runtime error do not delete
make_percent_change_map(
  file_name = "tk_movement_between_places_aggregated_admin.csv"
)
```

```{r, eval =F}
# Not working because of earlier runtime error do not delete
make_percent_change_map(
  file_name = "tk_movement_between_places_aggregated_bing.csv"
)
```


## Time Series of Movement

```{r fig-time-series, fig.cap = "There are strong difference in population changes across different counties", output = T}
top_counties <- tk_fb_data_admin |>
  group_by(county_geoid, county_name) |>
  summarise(
    max_abs_change = max(abs(percent_change), na.rm = TRUE),
    .groups = "drop"
  ) |>
  slice_max(max_abs_change, n = 20) |>
  pull(county_geoid)

fb_time <- tk_fb_data_admin |>
  filter(county_geoid %in% top_counties) |>
  mutate(ds = as.Date(ds)) |>
  group_by(ds, county_geoid, county_name) |>
  summarise(
    mean_percent_change = mean(percent_change, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(fb_time, aes(x = ds, y = mean_percent_change)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1) +
  facet_wrap(~ county_name, scales = "free_y") +
  scale_y_continuous() +
  labs(
    title = "% Population Density Change Over Time",
    x = NULL,
    y = "% Change vs Baseline"
  ) +
  theme_minimal()
```

## Places

As shown by @tbl-summary-table, ...

```{r tbl-summary-table, tbl.cap = "Bottom 20 county-to-county crisis movement flows", output = T}
county_flows <- tk_mp_data_admin |>
  group_by(start_county_name, end_county_name) |>
  summarise(
    total_crisis_movement = sum(n_crisis, na.rm = TRUE),
    avg_z_score = mean(z_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_crisis_movement))

county_flows |>
  slice_tail(n = 20) |>
  gt()

# Top 20 wasn't very interesting. Bottom 20 says something more informative 
```




