---
title: "Winter Storm Fern Situation Report - January 2026"
execute: 
  warning: false
  message: false
  output: false
  echo: false
fig-pos: H
format:
  pdf:
    colorlinks: true
    pdf-engine: xelatex
---


Formatted based on LA Fire [Situation Report](https://www.crisisready.io/wp-content/uploads/2025/01/ReadyMapper-LAFires-01152025.pdf)

# Summary

A major American winter storm, often referred to as Winter Storm Fern, started on Friday, January 23 and continued through January 26th. The storm brought heavy snow and freezing rain to several U.S. states, ranging from the southern plains to the Eastern United States. Several states experienced network outages and extremely cold temperatures.

The following visualizations are developed based on Meta AI for Good's data through Facebook. The data covers network coverage and population density from January 27 to February 10, with a focus on Tennessee and Kentucky and daily updates during the collection period.


# Movement

```{r}
library(ggplot2)
library(gt)
library(scales)
library(lubridate)
library(ggmap)
library(readr)
library (here)
library(dplyr)
library(sf)
library(tigris)
```

```{r}
make_percent_change_map <- function(file_name,
                                    map_date = NULL,
                                    date_col = "ds",
                                    value_col = "percent_change",
                                    output_plot_name = NULL,
                                    title_prefix = "Population Density Change on",
                                    bbox_buffer = 0.25) {
  
  # Read file from root Outputs folder
  df <- read_csv(
    here("Outputs", file_name),
    show_col_types = FALSE
  )
  
  # Make sure the date column is treated as a Date
  df <- df |>
    mutate("{date_col}" := as.Date(.data[[date_col]]))
  
  # If no date is supplied, use the latest date by default
  if (is.null(map_date)) {
    map_date <- max(df[[date_col]], na.rm = TRUE)
  } else {
    map_date <- as.Date(map_date)
  }
  
  # Summarize to county level for the selected date
  selected_county <- df |>
    filter(.data[[date_col]] == map_date) |>
    group_by(county_geoid, county_name, county_state) |>
    summarise(
      mean_percent_change = mean(.data[[value_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  if (nrow(selected_county) == 0) {
    stop("No data found for selected date: ", map_date)
  }
  
  # Join county summaries to county polygons
  map_data <- counties_sf |>
    inner_join(selected_county, by = "county_geoid") |>
    st_transform(4326)
  
  # Build state boundary layer from states represented in the data
  state_lines <- tigris::states(cb = TRUE, year = 2023) |>
    st_transform(4326) |>
    filter(NAME %in% unique(map_data$county_state))
  
  # Build bounding box around relevant counties
  bbox <- st_bbox(state_lines)
  
  xlim <- c(
    as.numeric(bbox["xmin"]) - bbox_buffer,
    as.numeric(bbox["xmax"]) + bbox_buffer
  )
  
  ylim <- c(
    as.numeric(bbox["ymin"]) - bbox_buffer,
    as.numeric(bbox["ymax"]) + bbox_buffer
  )
  
  # Build map using sf only
  p <- ggplot() +
    geom_sf(
      data = map_data,
      aes(fill = mean_percent_change),
      color = "black",
      linewidth = 0.25
    ) +
    geom_sf(
      data = state_lines,
      fill = NA,
      color = "black",
      linewidth = 0.7
    ) +
    scale_fill_gradient2(
      labels = label_number(suffix = "%", accuracy = 1)
    ) +
    coord_sf(
      xlim = xlim,
      ylim = ylim,
      expand = FALSE
    ) +
    labs(
      title = paste(title_prefix, map_date),
      fill = "% Change vs Baseline"
    ) +
    theme_minimal()
  
  # Save plot if requested
  if (!is.null(output_plot_name)) {
    ggsave(
      filename = here("Outputs", output_plot_name),
      plot = p,
      width = 10,
      height = 7
    )
  }
  
  return(p)
}
```



```{r}

# Get US county boundaries
options(tigris_use_cache = TRUE)
counties_sf <- counties(cb = TRUE, year = 2023) |>
  st_transform(4326) |>
  select(
    county_state = STATE_NAME,
    county_name = NAME,
    county_geoid = GEOID,
    geometry
  )
```

## WIP (Dont delete just hide)
```{r}
library(ggmap)
library(ggplot2)
library(readr)
library(dplyr)
library(here)
library(scales)
library(tigris)
library(sf)

options(tigris_use_cache = TRUE)
```

```{r}
# County boundaries
counties_sf <- tigris::counties(cb = TRUE, year = 2023) |>
  st_transform(4326) |>
  select(
    county_state = STATE_NAME,
    county_name = NAME,
    county_geoid = GEOID,
    geometry
  )

# State boundaries
states_sf <- tigris::states(cb = TRUE, year = 2023) |>
  st_transform(4326) |>
  select(
    state_name = NAME,
    geometry
  )
```


```{r}
sf_to_polygon_df <- function(sf_object, id_cols) {
  coords <- st_coordinates(sf_object)
  
  sf_object |>
    st_drop_geometry() |>
    mutate(.row_id = row_number()) |>
    right_join(
      as.data.frame(coords) |>
        mutate(.row_id = L1),
      by = ".row_id"
    ) |>
    mutate(
      group_id = interaction(.row_id, L1, L2, drop = TRUE)
    )
}
```

```{r}
make_percent_change_map <- function(file_name,
                                    map_date = NULL,
                                    date_col = "ds",
                                    value_col = "percent_change",
                                    output_plot_name = NULL,
                                    title_prefix = "Population Density Change on",
                                    bbox_buffer = 0.25,
                                    maptype = "stamen_toner_lite",
                                    zoom = 7) {
  
  # Read file from root Outputs folder
  df <- readr::read_csv(
    here::here("Outputs", file_name),
    show_col_types = FALSE
  )
  
  # Make sure the date column is treated as a Date
  df <- df |>
    dplyr::mutate("{date_col}" := as.Date(.data[[date_col]]))
  
  # If no date is supplied, use the latest date by default
  if (is.null(map_date)) {
    map_date <- max(df[[date_col]], na.rm = TRUE)
  } else {
    map_date <- as.Date(map_date)
  }
  
  # Summarize to county level for the selected date
  selected_county <- df |>
    dplyr::filter(.data[[date_col]] == map_date) |>
    dplyr::group_by(county_geoid, county_name, county_state) |>
    dplyr::summarise(
      mean_percent_change = mean(.data[[value_col]], na.rm = TRUE),
      .groups = "drop"
    )
  
  if (nrow(selected_county) == 0) {
    stop("No data found for selected date: ", map_date)
  }
  
  # Join selected data to county boundaries
  map_counties_sf <- counties_sf |>
    dplyr::inner_join(selected_county, by = "county_geoid")
  
  # Keep only the state boundaries represented in the data
  map_states_sf <- states_sf |>
    dplyr::filter(state_name %in% unique(map_counties_sf$county_state))
  
  # Use state boundaries to set the full map extent
  
 bbox <- sf::st_bbox(map_states_sf)

  bbox_map <- c(
    left   = as.numeric(unname(bbox["xmin"])) - bbox_buffer,
    bottom = as.numeric(unname(bbox["ymin"])) - bbox_buffer,
    right  = as.numeric(unname(bbox["xmax"])) + bbox_buffer,
    top    = as.numeric(unname(bbox["ymax"])) + bbox_buffer
  )
  
  names(bbox_map) <- c("left", "bottom", "right", "top")
  
  base_map <- ggmap::get_stadiamap(
    bbox = bbox_map,
    zoom = zoom,
    maptype = maptype
  )
  
  # Download ggmap/Stadia basemap
  base_map <- ggmap::get_stadiamap(
    bbox = bbox_map,
    zoom = zoom,
    maptype = maptype
  )
  
  # Convert county and state boundaries to regular data frames
  county_poly <- sf_to_polygon_df(map_counties_sf, id_cols = "county_geoid")
  
  state_poly <- sf_to_polygon_df(map_states_sf, id_cols = "state_name")
  
  # Build map using ggmap + regular polygon layers
  p <- ggmap::ggmap(base_map) +
    ggplot2::geom_polygon(
      data = county_poly,
      aes(
        x = X,
        y = Y,
        group = group_id,
        fill = mean_percent_change
      ),
      color = "white",
      linewidth = 0.25,
      alpha = 0.75,
      inherit.aes = FALSE
    ) +
    ggplot2::geom_path(
      data = state_poly,
      aes(
        x = X,
        y = Y,
        group = group_id
      ),
      color = "black",
      linewidth = 0.8,
      inherit.aes = FALSE
    ) +
    ggplot2::scale_fill_gradient2(
      labels = scales::label_number(suffix = "%", accuracy = 1)
    ) +
    ggplot2::labs(
      title = paste(title_prefix, map_date),
      fill = "% Change vs Baseline"
    ) +
    ggplot2::theme_minimal()
  
  # Save plot if requested
  if (!is.null(output_plot_name)) {
    ggplot2::ggsave(
      filename = here::here("Outputs", output_plot_name),
      plot = p,
      width = 10,
      height = 7
    )
  }
  
  return(p)
}
```


```{r}
make_percent_change_map(
  file_name = "tk_facebook_pop_aggregated_bing.csv",
  zoom = 7,
  bbox_buffer = 0.5
)
```



## % Population Density Change (Maps)

Note the difference in @fig-pop-bing, which is based on Bing tiles, and @fig-pop-admin, which is based on administrative regions.

```{r}
# Not working because of earlier runtime error do not delete
make_percent_change_map(
  file_name = "movement_between_places_aggregated.csv"
)
```

```{r fig-pop-bing, fig.cap = "Based on the Bing tiles, there are only a few counties with strong population density changes.", output = T}
make_percent_change_map(
  file_name = "tk_facebook_pop_aggregated_bing.csv",
   bbox_buffer = 1
)
```



```{r, eval=F}
# Not working because of earlier runtime error do not delete
make_percent_change_map(
  file_name = "tk_movement_between_places_aggregated_admin.csv"
)
```

```{r, eval =F}
# Not working because of earlier runtime error do not delete
make_percent_change_map(
  file_name = "tk_movement_between_places_aggregated_bing.csv"
)
```


## Time Series of Movement


```{r}
# Summarize county-level movement/change intensity 
# Have to decide between median and mean
county_change_summary <- tk_fb_data_admin |>
  mutate(ds = as.Date(ds)) |>
  group_by(county_geoid, county_name) |>
  summarise(
    avg_abs_change = mean(abs(percent_change), na.rm = TRUE),
    max_abs_change = max(abs(percent_change), na.rm = TRUE),
    .groups = "drop"
  )

# Select counties with highest, lowest, and average change
high_counties <- county_change_summary |>
  slice_max(avg_abs_change, n = 3) |>
  mutate(change_group = "Highest change")

low_counties <- county_change_summary |>
  slice_min(avg_abs_change, n = 3) |>
  mutate(change_group = "Lowest change")

average_value <- median(county_change_summary$avg_abs_change, na.rm = TRUE)

average_counties <- county_change_summary |>
  mutate(distance_from_average = abs(avg_abs_change - average_value)) |>
  slice_min(distance_from_average, n = 3) |>
  mutate(change_group = "Average change")
```

```{r}
# TS Prep function
make_county_time_data <- function(selected_counties) {
  tk_fb_data_admin |>
    mutate(ds = as.Date(ds)) |>
    inner_join(
      selected_counties |> select(county_geoid, change_group),
      by = "county_geoid"
    ) |>
    group_by(ds, county_geoid, county_name, change_group) |>
    summarise(
      mean_percent_change = mean(percent_change, na.rm = TRUE),
      median_percent_change = median(percent_change, na.rm = TRUE),
      .groups = "drop"
    )
}

# Plotting functions
plot_county_time_series <- function(time_data, plot_title) {
  ggplot(time_data, aes(x = ds, y = mean_percent_change, color = county_name)) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1) +
    scale_y_continuous(
      labels = scales::label_number(suffix = "%", accuracy = 1)
    ) +
    labs(
      title = plot_title,
      x = NULL,
      y = "% Change vs Baseline",
      color = "County"
    ) +
    theme_minimal()
}

## Facetted Option
plot_county_time_series_faceted <- function(time_data, plot_title) {
  ggplot(time_data, aes(x = ds, y = mean_percent_change, group = county_name)) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1) +
    facet_wrap(~ county_name, scales = "free_y") +
    scale_y_continuous(
      labels = scales::label_number(suffix = "%", accuracy = 1)
    ) +
    labs(
      title = plot_title,
      x = NULL,
      y = "% Change vs Baseline"
    ) +
    theme_minimal()
}

```


```{r fig-time-series, fig.cap = "There are strong difference in population changes across different counties", output = T}
high_time <- make_county_time_data(high_counties)

plot_county_time_series(
  high_time,
  "% Population Density Change Over Time: Highest-Change Counties"
)


average_time <- make_county_time_data(average_counties)

plot_county_time_series(
  average_time,
  "% Population Density Change Over Time: Average-Change Counties"
)


low_time <- make_county_time_data(low_counties)

plot_county_time_series(
  low_time,
  "% Population Density Change Over Time: Lowest-Change Counties"
)

```
Check why lowest tracks so closely on Feb 9th and investigate whether outliers are skewing mean
```{r fig-time-series, fig.cap = "There are strong difference in population changes across different counties", output = T}
high_time <- make_county_time_data(high_counties)

plot_county_time_series_faceted(
  high_time,
  "% Population Density Change Over Time: Highest-Change Counties"
)


average_time <- make_county_time_data(average_counties)

plot_county_time_series_faceted(
  average_time,
  "% Population Density Change Over Time: Average-Change Counties"
)


low_time <- make_county_time_data(low_counties)

plot_county_time_series_faceted(
  low_time,
  "% Population Density Change Over Time: Lowest-Change Counties"
)

```

```{r}
#-----------Investigation of mean v median---------------
# Might not even use but want thoughts on which to use, not many outliers present but there will be a lot of 0s pulling mean down
# Facetted plotting function using median percent change
plot_county_time_series_faceted_median <- function(time_data, plot_title) {
  ggplot(time_data, aes(x = ds, y = median_percent_change, group = county_name)) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1) +
    facet_wrap(~ county_name, scales = "free_y") +
    scale_y_continuous(
      labels = scales::label_number(suffix = "%", accuracy = 1)
    ) +
    labs(
      title = plot_title,
      subtitle = "County-level median percent change from baseline",
      x = NULL,
      y = "Median % Change vs Baseline"
    ) +
    theme_minimal()
}

high_time <- make_county_time_data(high_counties)

plot_county_time_series_faceted_median(
  high_time,
  "% Population Density Change Over Time: Highest-Change Counties"
)


average_time <- make_county_time_data(average_counties)

plot_county_time_series_faceted_median(
  average_time,
  "% Population Density Change Over Time: Average-Change Counties"
)


low_time <- make_county_time_data(low_counties)

plot_county_time_series_faceted_median(
  low_time,
  "% Population Density Change Over Time: Lowest-Change Counties"
)
```


## Places

As shown by @tbl-summary-table, ...

```{r tbl-summary-table, tbl.cap = "Bottom 20 county-to-county crisis movement flows", output = T}
county_flows <- tk_mp_data_admin |>
  group_by(start_county_name, end_county_name) |>
  summarise(
    total_crisis_movement = sum(n_crisis, na.rm = TRUE),
    avg_z_score = mean(z_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_crisis_movement))

county_flows |>
  slice_tail(n = 20) |>
  gt()

# Top 20 wasn't very interesting. Bottom 20 says something more informative 
```


# Demographics




