---
title: "Winter Storm Fern Situation Report"
subtitle: "January 2026"
execute: 
  warning: false
  message: false
  output: false
  echo: false
fig-pos: H
bibliography: references.bib
format:
  pdf:
    colorlinks: true
    pdf-engine: xelatex
    tbl-cap-location: bottom
    include-in-header:
      text: |
        \usepackage{etoolbox}

        \makeatletter
        \patchcmd{\@makefntext}
          {\parindent 1em}
          {\parindent 1em\leftskip -0.5in\rightskip -0.5in}
          {}{}
        \makeatother
---

Formatted based on LA Fire [Situation Report](https://www.crisisready.io/wp-content/uploads/2025/01/ReadyMapper-LAFires-01152025.pdf)

```{r}
library(ggplot2)
library(gt)
library(scales)
library(lubridate)
library(ggmap)
library(readr)
library(patchwork)
library (here)
library(dplyr)
library(sf)
library(tigris)
```

```{r}
# Get US county boundaries
options(tigris_use_cache = TRUE)
counties_sf <- counties(cb = TRUE, year = 2023) |>
  st_transform(4326) |>
  select(
    county_state = STATE_NAME,
    county_name = NAME,
    county_geoid = GEOID,
    geometry
  )

# State boundaries
states_sf <- tigris::states(cb = TRUE, year = 2023) |>
  st_transform(4326) |>
  select(
    state_name = NAME,
    geometry
  )
library(tidyverse)
library(sf)
library(patchwork)  # For arranging plots
library(scales)     # For squish

make_percent_change_map <- function(file_name,
                                    date_col = "ds",
                                    value_col = "percent_change",
                                    selected_date = NULL,   # Can be length 1 or 2
                                    output_plot_name = NULL,
                                    title_prefix = "Population Density Change (%)") {
  
  # Read file
  df <- read_csv(
    here("outputs", file_name),
    show_col_types = FALSE
  )
  
  # Determine which dates to use
  if (is.null(selected_date)) {
    plot_dates <- max(df[[date_col]], na.rm = TRUE)
  } else {
    if (!all(selected_date %in% df[[date_col]])) {
      stop("One or more selected_date values are not available in the data.")
    }
    plot_dates <- selected_date
  }
  
 # Compute global fill range using the summarized county-level data
  all_summarized <- df |>
    filter(.data[[date_col]] %in% plot_dates) |>
    group_by(county_geoid, county_name, county_state, !!sym(date_col)) |>
    summarise(mean_percent_change = mean(.data[[value_col]], na.rm = TRUE), .groups = "drop")
  
  fill_range <- all_summarized |>
    summarise(min_val = min(mean_percent_change, na.rm = TRUE),
              max_val = max(mean_percent_change, na.rm = TRUE))
  
  # Function to create a single map
  create_map <- function(plot_date) {
    latest_county <- df |>
      filter(.data[[date_col]] == plot_date) |>
      group_by(county_geoid, county_name, county_state) |>
      summarise(
        mean_percent_change = mean(.data[[value_col]], na.rm = TRUE),
        .groups = "drop"
      )
    
    map_data <- counties_sf |>
      inner_join(latest_county, by = "county_geoid")
    
    county_bbox <- st_bbox(map_data)
    relevant_states <- states_sf |>
      st_filter(map_data, .predicate = st_intersects)
    
    ggplot() +
      geom_sf(data = map_data, aes(fill = mean_percent_change), linewidth = 0.2) +
      geom_sf(data = relevant_states, fill = NA, color = "black", linewidth = 0.7) +
      scale_fill_gradient2(
        limits = c(fill_range$min_val, fill_range$max_val),
        oob = squish,   # ensure values outside the plot's range are included
        labels = label_number(suffix = "%", accuracy = 1)
      ) +
      labs(
        title = plot_date,
        fill = "% Change vs Baseline"
      ) +
      coord_sf(
        xlim = c(county_bbox["xmin"], county_bbox["xmax"]),
        ylim = c(county_bbox["ymin"], county_bbox["ymax"])
      ) +
      theme_minimal()
  }
  
  # Create plots
  plots <- lapply(plot_dates, create_map)
  
  # Combine plots side-by-side with shared legend
  if (length(plots) == 1) {
    combined_plot <- plots[[1]] +
      ggtitle(paste(title_prefix, plots[[1]]$labels$title))
  } else {
    combined_plot <- wrap_plots(plots, ncol = length(plots), guides = "collect") &
      theme(legend.position = "bottom")
    
    # Shared title
    combined_plot <- combined_plot + plot_annotation(
    title = title_prefix,
    theme = theme(
      plot.title = element_text(hjust = 0.5)  # centers the title
    )
  )
  }
  
  # Save if requested
  if (!is.null(output_plot_name)) {
    ggsave(
      filename = here("outputs", output_plot_name),
      plot = combined_plot,
      width = ifelse(length(plot_dates) == 2, 16, 10),
      height = 7
    )
  }
  
  return(combined_plot)
}
```


```{r}
bing_vs_admin_map <- function(date_col = "ds",
                                    value_col = "percent_change",
                                    selected_date = NULL,   # Can be length 1 or 2
                                    output_plot_name = NULL,
                                    title_prefix = "Population Density Change (%)") {
  
  # Read file
  df_bing <- read_csv(
    here("outputs", "tk_facebook_pop_aggregated_bing.csv"),
    show_col_types = FALSE
  )
  
  df_admin <- read_csv(
    here("outputs", "tk_facebook_pop_aggregated_admin.csv"),
    show_col_types = FALSE
  )
  
  # Determine which dates to use
  # Combine first so date checks use all available data
  common_cols <- intersect(names(df_bing), names(df_admin))
  df_combined <- rbind(df_bing[, common_cols], df_admin[, common_cols])
  
  # Determine which dates to use
  if (is.null(selected_date)) {
    plot_dates <- max(df_combined[[date_col]], na.rm = TRUE)
  } else {
    if (!all(selected_date %in% df_combined[[date_col]])) {
      stop("One or more selected_date values are not available in the data.")
    }
    plot_dates <- selected_date
  }
  
 # Compute global fill range using the summarized county-level data
  all_summarized <- df_combined |>
    filter(.data[[date_col]] %in% plot_dates) |>
    group_by(county_geoid, county_name, county_state, !!sym(date_col)) |>
    summarise(mean_percent_change = mean(.data[[value_col]], na.rm = TRUE), .groups = "drop")
  
  fill_range <- all_summarized |>
    summarise(min_val = min(mean_percent_change, na.rm = TRUE),
              max_val = max(mean_percent_change, na.rm = TRUE))
  
  # Function to create a single map
  create_map <- function(df, source, plot_date = plot_dates) {

    latest_county <- df |>
      filter(.data[[date_col]] == plot_date) |>
      group_by(county_geoid, county_name, county_state) |>
      summarise(
        mean_percent_change = mean(.data[[value_col]], na.rm = TRUE),
        .groups = "drop"
      )
  
    map_data <- counties_sf |>
      inner_join(latest_county, by = "county_geoid")
  
    county_bbox <- st_bbox(map_data)
  
    relevant_states <- states_sf |>
      st_filter(map_data, .predicate = st_intersects)
  
    ggplot() +
      geom_sf(
        data = map_data,
        aes(fill = mean_percent_change),
        linewidth = 0.2
      ) +
      geom_sf(
        data = relevant_states,
        fill = NA,
        color = "black",
        linewidth = 0.7
      ) +
      scale_fill_gradient2(
        limits = c(fill_range$min_val, fill_range$max_val),
        oob = squish,
        labels = label_number(suffix = "%", accuracy = 1)
      ) +
      labs(
        title = paste0(plot_date, " (", source, ")"),
        fill = "% Change vs Baseline"
      ) +
      coord_sf(
        xlim = c(county_bbox["xmin"], county_bbox["xmax"]),
        ylim = c(county_bbox["ymin"], county_bbox["ymax"])
      ) +
      theme_minimal()
  }
  
  # Create plots
  plots <- list(
  create_map(df_bing, source = "Bing"),
  create_map(df_admin, source = "Admin")
)
  
  # Combine plots side-by-side with shared legend
  if (length(plots) == 1) {
    combined_plot <- plots[[1]] +
      ggtitle(paste(title_prefix, plots[[1]]$labels$title))
  } else {
    combined_plot <- wrap_plots(plots, ncol = length(plots), guides = "collect") &
      theme(legend.position = "bottom")
    
    # Shared title
    combined_plot <- combined_plot + plot_annotation(
    title = title_prefix,
    theme = theme(
      plot.title = element_text(hjust = 0.5)  # centers the title
    )
  )
  }
  
  # Save if requested
  if (!is.null(output_plot_name)) {
    ggsave(
      filename = here("outputs", output_plot_name),
      plot = combined_plot,
      width = ifelse(length(plot_dates) == 2, 16, 10),
      height = 7
    )
  }
  
  return(combined_plot)
}
```


# Summary

A major American winter storm, often referred to as Winter Storm Fern, started on Friday, January 23rd 2026 and continued through January 26th 2026. The storm brought heavy snow and freezing rain to several U.S. states, ranging from the southern plains to the Eastern United States. Several states experienced network outages and extremely cold temperatures. For example, emergency declarations were issued in Arkansas, Georgia, Indiana, Kentucky, Louisiana, Maryland, Mississippi, North Carolina, South Carolina, Tennessee, Virginia, and West Virginia [@crs_in12644].

The following visualizations are developed based on Meta AI for Good's data through Facebook with a focus on Tennessee and Kentucky.. The data covers network coverage and population density from January 27 to February 10, with a focus on Tennessee and Kentucky and daily updates during the collection period.

# How the Data Is Constructed

The Facebook data used in this report comes from Meta’s Data for Good crisis datasets. In plain terms, these datasets use activity from Facebook users or Facebook business Pages to estimate how people, population density, network coverage, and business activity change during a crisis. The data does not represent everyone in the affected area. For population and movement datasets, it only represents Facebook app users who have Location Services enabled. For business activity, it represents qualifying Facebook business Pages with enough activity to be included while preserving privacy. 

The data is constructed by comparing what is observed during the crisis period to what was typical before the crisis. For example, population density during the storm is compared to a pre-crisis baseline. Movement during the storm is compared to normal movement patterns between the same places before the event. Business activity is compared to normal posting behavior from business pages before the crisis. Because the data is relative to a baseline, the main signal is not the raw count itself, but whether a place is above or below its expected level.

Geographically, the data is reported in two main ways: Bing tiles and administrative regions. Bing tiles divide space into small grid cells. At Bing tile level 14, each tile is roughly 2.4 kilometers on a side near the equator, or about the size of several city blocks. Administrative regions instead use boundaries such as counties, states, or other official geographic areas. These are two different approximations of the same general region and same time window, but they can show different rates of population density change ( @fig-bing-vs-admin).

```{r fig-bing-vs-admin, fig.cap = "Changes in population density vary when represented by data from bing tiles or by data from administrative regions, even in the same region and time window.", output = T}
bing_vs_admin_map(selected_date = "2026-02-12")
```
**Takeaways** (@fig-bing-vs-admin)

- Bing tile results and administrative-region results should not be treated as interchangeable

- The strongest changes appear concentrated in a small number of counties rather than spread evenly across the region


Temporally, the population and movement datasets use fixed 8-hour windows. These windows begin at 00:00, 08:00, and 16:00 Pacific Time. This means that the time periods do not automatically adjust to local time zones. For Tennessee and Kentucky, the reported time windows are still based on Pacific Time, which should be kept in mind when interpreting daily patterns. Business activity is reported daily, based on the date value in the dataset. 
Some Meta crisis datasets are provided directly as Bing tile data, while others are provided as administrative-region data. When the data is provided as Bing tiles, each observation is tied to a grid cell with a latitude and longitude. When we add county information to those data, we are not changing the original measurement. Instead, we are assigning each tile or point to the county that contains it. This allows the tile-level data to be summarized at the county level.

For example, a Bing tile located inside Davidson County, Tennessee would be assigned the Davidson County name and GEOID. If many Bing tiles fall within Davidson County, their values can then be averaged, summed, or otherwise summarized to produce a county-level measure. This is useful for reporting, but it is not identical to receiving an administrative-region dataset directly from Meta. The administrative-region dataset is already aggregated within county or state boundaries by Meta, while the Bing tile version requires us to overlay the tile locations onto county boundaries.


# Facebook Data Definitions

- **Baseline Ratio**: Ratio of people using the Facebook app who have turned on the Location Services device setting on their mobile device that are present in a specific tile during the baseline period over users in all tiles in the boundary box defining the crisis

- **Crisis Ratio**: Ratio of people using the Facebook app who have turned on the Location Services device setting on their mobile device that are present in a specific tile during the 8-hour window over users in all tiles in the boundary box defining the crisis

- **Population Change Density (%)**: Percent change in Facebook Location Services-enabled user population counts per tile, calculated by dividing the difference by the baseline (plus a small value, usually 1)

- **Time Windows**: aggregated over 3 fixed 8-hour time windows, beginning at 00:00, 08:00 and 16:00, that represent changes in population density

- **Bing Tiles**: equivalent to roughly 2.4 kilometers on a side near the equator or the size of 6 x 6 city blocks

- **Administrative Regions**: the polygon boundaries for administrative regions derived from the Database of Global Administrative Areas, also known as GADM.



# Outside Data Added

In addition to the Meta crisis datasets, county-level demographic data was added from the American Community Survey [@acs]. This demographic data was incorporated with the use of the `tidycensus` package [@tidycensus]. This outside data provides context about the counties affected by the storm, including total population, median household income, poverty rate, share of residents age 65 and older, and share of households without vehicle access. These variables help interpret which communities may be more vulnerable during a crisis.

The demographic data is joined using `county_geoid`, a standardized county identifier. This is important because county names alone are not unique across states. For example, many states have counties with the same name, but each county has a unique GEOID.

# Aggregation Choices

For county-level maps, the data is summarized by county and date. When multiple observations fall within the same county on the same date, the report uses the average percent change to represent the county. This choice is appropriate for percent-based indicators because summing percentages would not be meaningful.

For population count variables, such as baseline population or crisis population, summing can be used when the goal is to estimate total observed users across tiles or regions. For percent change, z-scores, or ratios, averaging is more appropriate because these variables already describe relative change.

For movement data, origin and destination counties are treated separately. Each row represents movement from a starting location to an ending location, so the analysis keeps both `start_county_geoid` and `end_county_geoid`. This avoids incorrectly collapsing movement into a single county and preserves the direction of movement.

# Limitations

These datasets should be interpreted as indicators of crisis-related change, not exact measures of the full population. The population and movement data only include Facebook users with Location Services enabled, so the sample may not represent all residents equally. Areas with fewer Facebook users or lower Location Services usage may be less reliable.

Privacy protections also affect interpretation. Small counts may be suppressed or removed, which can especially affect rural areas. For movement data, this means it is not reliable to calculate exact net flows by simply adding inbound and outbound movement, because some small movement vectors are excluded. 

The baseline period is limited to the pre-crisis period available to Meta. This means the analysis does not have a long historical record of pre-crisis behavior. As a result, seasonal changes, holidays, or unusual pre-crisis conditions may affect the baseline. The data is useful for rapid situational awareness, but it should not be interpreted as a complete long-term population or economic dataset.

Finally, the data is time-limited. The documentation notes that some datasets are accessible for 90 days after the final update day and then discarded, which limits longer-term reproducibility and longitudinal analysis.


{{<pagebreak>}}

# Movement

## % Population Density Change


```{r fig-pop-bing, fig.cap = "When showing population density change with Bing tiles, there is little negative changes at these two dates.", output = T}
make_percent_change_map(
  file_name = "tk_facebook_pop_aggregated_bing.csv",
  selected_date = c("2026-02-02", "2026-02-12"),
  title_prefix = "Population Density Change (%) - Bing Tiles"
)
```

**Takeaways** (@fig-pop-bing)

- The overall spatial pattern appears fairly stable between the two dates

- There is little visible negative population-density change in the Bing tile version

```{r fig-pop-admin, fig.cap = "When showing population density change with administrative regions, there are a few counties with very strong negative change during both dates.", output = T}
make_percent_change_map(
  file_name = "tk_facebook_pop_aggregated_admin.csv",
  selected_date = c("2026-02-02", "2026-02-12"),
  title_prefix = "Population Density Change (%) - Admin Regions"
)
```
**Takeaways** (@fig-pop-admin)

- The administrative-region maps show more visible negative population-density change than the Bing tile maps.

- A few counties stand out with strong decreases on both February 2 and February 12.

- The contrast with Figure 2 shows that the same crisis can look different depending on whether the data is summarized by tiles or administrative regions.

## Time Series of Movement


```{r}
tk_fb_data_admin <- read_csv(here("Outputs", "tk_facebook_pop_aggregated_bing.csv"), show_col_types = F)
# Summarize county-level movement/change intensity 
# Have to decide between median and mean
county_change_summary <- tk_fb_data_admin |>
  mutate(ds = as.Date(ds)) |>
  group_by(county_geoid, county_name) |>
  summarise(
    avg_abs_change = mean(abs(percent_change), na.rm = TRUE),
    max_abs_change = max(abs(percent_change), na.rm = TRUE),
    .groups = "drop"
  )

# Select counties with highest, lowest, and average change
high_counties <- county_change_summary |>
  slice_max(avg_abs_change, n = 3) |>
  mutate(change_group = "Highest change")

low_counties <- county_change_summary |>
  slice_min(avg_abs_change, n = 3) |>
  mutate(change_group = "Lowest change")

average_value <- median(county_change_summary$avg_abs_change, na.rm = TRUE)

average_counties <- county_change_summary |>
  mutate(distance_from_average = abs(avg_abs_change - average_value)) |>
  slice_min(distance_from_average, n = 3) |>
  mutate(change_group = "Average change")
```


```{r}
# TS Prep function
make_county_time_data <- function(selected_counties) {
  tk_fb_data_admin |>
    mutate(ds = as.Date(ds)) |>
    inner_join(
      selected_counties |> select(county_geoid, change_group),
      by = "county_geoid"
    ) |>
    group_by(ds, county_geoid, county_name, change_group) |>
    summarise(
      mean_percent_change = mean(percent_change, na.rm = TRUE),
      median_percent_change = median(percent_change, na.rm = TRUE),
      .groups = "drop"
    )
}

# Plotting functions
plot_county_time_series <- function(time_data, plot_title) {
  ggplot(time_data, aes(x = ds, y = mean_percent_change, color = county_name)) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1) +
    scale_y_continuous(
      labels = scales::label_number(suffix = "%", accuracy = 1)
    ) +
    labs(
      title = plot_title,
      x = NULL,
      y = "% Change vs Baseline",
      color = "County"
    ) +
    theme_minimal()
}

## Facetted Option
plot_county_time_series_faceted <- function(time_data, plot_title) {
  ggplot(time_data, aes(x = ds, y = mean_percent_change, group = county_name)) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1) +
    facet_wrap(~ county_name, scales = "free_y") +
    scale_y_continuous(
      labels = scales::label_number(suffix = "%", accuracy = 1)
    ) +
    labs(
      title = plot_title,
      x = NULL,
      y = "% Change vs Baseline"
    ) +
    theme_minimal() + 
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

}

```
The counties are grouped by average absolute population-density change over the recording period.[^avg-change]

[^avg-change]: The percent changes are summarized over the full recording period for each county. For each county, the absolute value of the percent change is taken at each time point, and those absolute changes are averaged across the 14-day period. This creates a single measure of how much each county deviated from baseline, regardless of whether the change was positive or negative. Counties are then compared using this average absolute change to identify the highest-change, lowest-change, and average-change counties. The county-level change measure is defined as $\text{AvgAbsChange}_i = \frac{1}{T_i}\sum_{t=1}^{T_i} |p_{it}|$, where $p_{it}$ is the percent change for county $i$ at time $t$, and $T_i$ is the number of time periods observed for that county.

```{r fig-TS-high, fig.cap = "The three counties with the highest change have positive change or negative change during this entire time period.", output = T}
high_time <- make_county_time_data(high_counties)

plot_county_time_series(
  high_time,
  "% Population Density Change Over Time:\nHighest-Change Counties"
)
```
**Takeaways** (@fig-TS-high)

- Fairfield and Issaquena show the largest positive population-density changes over the time period.

- Issaquena rises sharply near the end of the period, reaching close to 100% change.

- Sevier is included among the highest-change counties because of large absolute change, but its trend is negative rather than positive.

- The figure shows that “highest change” includes both large increases and large decreases when counties are selected using absolute change.

- The time series makes clear that the largest changes are persistent over multiple days, not just one-day spikes.

```{r fig-TS-medium, fig.cap = "The three counties with average change have more variation in population density change with both positive and negative changes.", output = T}
average_time <- make_county_time_data(average_counties)

plot_county_time_series(
  average_time,
  "% Population Density Change Over Time:\nAverage-Change Counties"
)
```
**Takeaways** (@fig-TS-medium)

- These counties provide a useful comparison group for interpreting the more extreme counties.


```{r fig-TS-low, fig.cap = "The three counties with the lowest change have a notable drop into extreme negative population density change on February 9th.",  output = T}
low_time <- make_county_time_data(low_counties)

plot_county_time_series(
  low_time,
  "% Population Density Change Over Time:\nLowest-Change Counties"
)

```
**Takeaways** (@fig-TS-low)

- I got nothing

{{<pagebreak>}}

## Places



```{r, output = T}
#| label: tbl-summary-table
#| tbl-cap: "The bottom 20 county-to-county crisis movement flows are all at zero or ten, likely due to data-collection thresholds for privacy concerns."
tk_mp_data_bing <- read_csv(here("Outputs", "tk_movement_between_places_aggregated_bing.csv"),
                             show_col_types = F)
county_flows <- tk_mp_data_bing |>
  group_by(start_county_name, end_county_name) |>
  summarise(
    total_crisis_movement = sum(n_crisis, na.rm = TRUE),
    avg_z_score = mean(z_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(total_crisis_movement)

county_flows |>
  slice_head(n = 20) |>
  gt() |>
  tab_header(
    title = "Bottom 20 county-to-county crisis movement flows",
    subtitle = "Source: Bing Tiles"
  )

```
**Takeaways** (@tbl-summary-table)

- The table is more useful as a data-quality warning than as a substantive movement analysis.

- Because movement vectors with small counts can be suppressed, these low-flow values should not be interpreted as exact zero movement.

{{< pagebreak >}}

# Demographic Profile

```{r}

make_demographics_table <- function(file_name, 
                                    map_date = NULL,
                                    date_col = "ds",
                                    value_col = "percent_change",
                                    n_counties = 15,
                                    output_table_name = NULL,
                                    table_title = "Demographic Profile of Most Impacted Counties", desc = T,
                                    source) {
  
  # Read file from Outputs folder
  df <- read_csv(
    here("Outputs", file_name),
    show_col_types = FALSE
  )
  
  # Convert date column
  df <- df |>
    mutate("{date_col}" := as.Date(.data[[date_col]]))
  
  # Use latest date by default
  if (is.null(map_date)) {
    map_date <- max(df[[date_col]], na.rm = TRUE)
  } else {
    map_date <- as.Date(map_date)
  }
  
  # Summarize to county level
  county_demo <- df |>
    filter(.data[[date_col]] == map_date) |>
    group_by(
      county_geoid,
      county_name,
      county_state
    ) |>
    summarise(
      mean_percent_change = mean(.data[[value_col]], na.rm = TRUE),
      total_population = first(na.omit(total_population)),
      median_income = first(na.omit(median_income)),
      poverty_rate = first(na.omit(poverty_rate)),
      pct_age_65_plus = first(na.omit(pct_age_65_plus)),
      pct_no_vehicle = first(na.omit(pct_no_vehicle)),
      .groups = "drop"
    ) |>
    mutate(
      abs_change = abs(mean_percent_change)
    )
  
  # Create gt table
  demo_table <- county_demo |>
    select(
      County = county_name,
      State = county_state,
      `Population Density Change` = mean_percent_change,
      `Total Population` = total_population,
      `Median Household Income` = median_income,
      `Poverty Rate` = poverty_rate,
      `% Age 65+` = pct_age_65_plus,
      `% No Vehicle Access` = pct_no_vehicle
    )
  demo_table <- arrange(demo_table, `Population Density Change`)
  if (desc){
    demo_table <- arrange(demo_table, desc(`Population Density Change`))
  }
  demo_table |>
  head(n_counties) |>
  gt() |>
  tab_header(
    title = table_title,
    subtitle = paste0("Source: ", source)
  ) |>
  # Wrap columns
  cols_width(
    County ~ px(120),
    State ~ px(120),
    `Population Density Change` ~ px(70),
    `Total Population` ~ px(70),
    `Median Household Income` ~ px(70),
    everything() ~ px(50)
  ) |>
  tab_options(
    table.font.size = px(9),
    heading.title.font.size = px(16),
    heading.subtitle.font.size = px(9),
    data_row.padding = px(25),
    column_labels.padding = px(0),
    table.width = pct(100)
  ) |>
  fmt_number(
    columns = `Population Density Change`,
    decimals = 1,
    suffixing = FALSE
  ) |>
  fmt_number(
    columns = `Total Population`,
    decimals = 0,
    use_seps = TRUE
  ) |>
  fmt_currency(
    columns = `Median Household Income`,
    currency = "USD",
    decimals = 0
  ) |>
  fmt_percent(
    columns = c(`Poverty Rate`, `% Age 65+`, `% No Vehicle Access`),
    decimals = 1
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  opt_table_font(
    font = list(
      google_font("Roboto Condensed"),
      default_fonts()
    )
  )
}
```


```{r, output = T}
#| label: tbl-bing02-inc
#| tbl-cap: "There are a few counties with high population density change but low total population."

make_demographics_table(
  file_name = "tk_facebook_pop_aggregated_bing.csv",
  map_date = "2026-02-02",
  n_counties = 15,
  source = "Bing Tiles",
  table_title = "Counties with Most Increased Population Density on 2026-02-02"
) 
```

```{r, output = T}
#| label: tbl-bing02-dec
#| tbl-cap: "Most of the counties with the lowest population density change have fairly high vehicular access."

make_demographics_table(
  file_name = "tk_facebook_pop_aggregated_bing.csv",
  n_counties = 15,
  map_date = "2026-02-02",
  desc = F,
  source = "Bing Tiles",
  table_title = "Counties with Most Decreased Population Density on 2026-02-02"
) 
```

```{r, output = T}
#| label: tbl-bing12-inc
#| tbl-cap: "Issaquena has the highest population density but notably small population size."

make_demographics_table(
  file_name = "tk_facebook_pop_aggregated_bing.csv",
  map_date = "2026-02-12",
  n_counties = 15,
  source = "Bing Tiles",
  table_title = "Counties with Most Increased Population Density on 2026-02-12"
) 
```

```{r, output = T}
#| label: tbl-bing12-dec
#| tbl-cap: "Similar to Feb 2nd, the counties with strong negative population density change also have high levels of vehicular access."

make_demographics_table(
  file_name = "tk_facebook_pop_aggregated_bing.csv",
  n_counties = 15,
  map_date = "2026-02-12",
  desc = F,
  source = "Bing Tiles",
  table_title = "Counties with Most Decreased Population Density on 2026-02-12"
  ) 
```

{{< pagebreak >}}

# References