---
title: "Winter Storm Fern Situation Report - January 2026"
execute: 
  warning: false
  message: false
  output: false
  echo: false
fig-pos: H
format:
  pdf:
    colorlinks: true
    pdf-engine: xelatex
---


Formatted based on LA Fire [Situation Report](https://www.crisisready.io/wp-content/uploads/2025/01/ReadyMapper-LAFires-01152025.pdf)

# Summary

A major American winter storm, often referred to as Winter Storm Fern, started on Friday, January 23 and continued through January 26th. The storm brought heavy snow and freezing rain to several U.S. states, ranging from the southern plains to the Eastern United States. Several states experienced network outages and extremely cold temperatures.

The following visualizations are developed based on Meta AI for Good's data through Facebook. The data covers network coverage and population density from January 27 to February 10, with a focus on Tennessee and Kentucky and daily updates during the collection period.


# Movement

```{r}
library(ggplot2)
library(gt)
library(scales)
library(lubridate)
library(ggmap)
library(readr)
library(patchwork)
library (here)
library(dplyr)
library(sf)
library(tigris)
```

```{r}
# Get US county boundaries
options(tigris_use_cache = TRUE)
counties_sf <- counties(cb = TRUE, year = 2023) |>
  st_transform(4326) |>
  select(
    county_state = STATE_NAME,
    county_name = NAME,
    county_geoid = GEOID,
    geometry
  )

# State boundaries
states_sf <- tigris::states(cb = TRUE, year = 2023) |>
  st_transform(4326) |>
  select(
    state_name = NAME,
    geometry
  )
library(tidyverse)
library(sf)
library(patchwork)  # For arranging plots
library(scales)     # For squish

make_percent_change_map <- function(file_name,
                                    date_col = "ds",
                                    value_col = "percent_change",
                                    selected_date = NULL,   # Can be length 1 or 2
                                    output_plot_name = NULL,
                                    title_prefix = "Population Density Change (%)") {
  
  # Read file
  df <- read_csv(
    here("outputs", file_name),
    show_col_types = FALSE
  )
  
  # Determine which dates to use
  if (is.null(selected_date)) {
    plot_dates <- max(df[[date_col]], na.rm = TRUE)
  } else {
    if (!all(selected_date %in% df[[date_col]])) {
      stop("One or more selected_date values are not available in the data.")
    }
    plot_dates <- selected_date
  }
  
 # Compute global fill range using the summarized county-level data
  all_summarized <- df |>
    filter(.data[[date_col]] %in% plot_dates) |>
    group_by(county_geoid, county_name, county_state, !!sym(date_col)) |>
    summarise(mean_percent_change = mean(.data[[value_col]], na.rm = TRUE), .groups = "drop")
  
  fill_range <- all_summarized |>
    summarise(min_val = min(mean_percent_change, na.rm = TRUE),
              max_val = max(mean_percent_change, na.rm = TRUE))
  
  # Function to create a single map
  create_map <- function(plot_date) {
    latest_county <- df |>
      filter(.data[[date_col]] == plot_date) |>
      group_by(county_geoid, county_name, county_state) |>
      summarise(
        mean_percent_change = mean(.data[[value_col]], na.rm = TRUE),
        .groups = "drop"
      )
    
    map_data <- counties_sf |>
      inner_join(latest_county, by = "county_geoid")
    
    county_bbox <- st_bbox(map_data)
    relevant_states <- states_sf |>
      st_filter(map_data, .predicate = st_intersects)
    
    ggplot() +
      geom_sf(data = map_data, aes(fill = mean_percent_change), linewidth = 0.2) +
      geom_sf(data = relevant_states, fill = NA, color = "black", linewidth = 0.7) +
      scale_fill_gradient2(
        limits = c(fill_range$min_val, fill_range$max_val),
        oob = squish,   # ensure values outside the plot's range are included
        labels = label_number(suffix = "%", accuracy = 1)
      ) +
      labs(
        title = plot_date,
        fill = "% Change vs Baseline"
      ) +
      coord_sf(
        xlim = c(county_bbox["xmin"], county_bbox["xmax"]),
        ylim = c(county_bbox["ymin"], county_bbox["ymax"])
      ) +
      theme_minimal()
  }
  
  # Create plots
  plots <- lapply(plot_dates, create_map)
  
  # Combine plots side-by-side with shared legend
  if (length(plots) == 1) {
    combined_plot <- plots[[1]] +
      ggtitle(paste(title_prefix, plots[[1]]$labels$title))
  } else {
    combined_plot <- wrap_plots(plots, ncol = length(plots), guides = "collect") &
      theme(legend.position = "bottom")
    
    # Shared title
    combined_plot <- combined_plot + plot_annotation(
    title = title_prefix,
    theme = theme(
      plot.title = element_text(hjust = 0.5)  # centers the title
    )
  )
  }
  
  # Save if requested
  if (!is.null(output_plot_name)) {
    ggsave(
      filename = here("outputs", output_plot_name),
      plot = combined_plot,
      width = ifelse(length(plot_dates) == 2, 16, 10),
      height = 7
    )
  }
  
  return(combined_plot)
}
```

## % Population Density Change

Note the difference in @fig-pop-bing, which is based on Bing tiles, and @fig-pop-admin, which is based on administrative regions.

```{r fig-pop-bing, fig.cap = "Based on the Bing tiles, there are only a few counties with strong population density changes.", output = T}
make_percent_change_map(
  file_name = "tk_facebook_pop_aggregated_bing.csv",
  selected_date = c("2026-02-02", "2026-02-12")
)
```


```{r fig-pop-admin, fig.cap = "Based on the administrative regions, there are two counties with strong population density changes, one positively and one negatively.", output = T}
make_percent_change_map(
  file_name = "tk_facebook_pop_aggregated_admin.csv",
  selected_date = c("2026-01-30", "2026-02-12")
)
```


## Time Series of Movement


```{r}
tk_fb_data_admin <- read_csv(here("Outputs", "tk_facebook_pop_aggregated_bing.csv"), show_col_types = F)
# Summarize county-level movement/change intensity 
# Have to decide between median and mean
county_change_summary <- tk_fb_data_admin |>
  mutate(ds = as.Date(ds)) |>
  group_by(county_geoid, county_name) |>
  summarise(
    avg_abs_change = mean(abs(percent_change), na.rm = TRUE),
    max_abs_change = max(abs(percent_change), na.rm = TRUE),
    .groups = "drop"
  )

# Select counties with highest, lowest, and average change
high_counties <- county_change_summary |>
  slice_max(avg_abs_change, n = 3) |>
  mutate(change_group = "Highest change")

low_counties <- county_change_summary |>
  slice_min(avg_abs_change, n = 3) |>
  mutate(change_group = "Lowest change")

average_value <- median(county_change_summary$avg_abs_change, na.rm = TRUE)

average_counties <- county_change_summary |>
  mutate(distance_from_average = abs(avg_abs_change - average_value)) |>
  slice_min(distance_from_average, n = 3) |>
  mutate(change_group = "Average change")
```

```{r}
# TS Prep function
make_county_time_data <- function(selected_counties) {
  tk_fb_data_admin |>
    mutate(ds = as.Date(ds)) |>
    inner_join(
      selected_counties |> select(county_geoid, change_group),
      by = "county_geoid"
    ) |>
    group_by(ds, county_geoid, county_name, change_group) |>
    summarise(
      mean_percent_change = mean(percent_change, na.rm = TRUE),
      median_percent_change = median(percent_change, na.rm = TRUE),
      .groups = "drop"
    )
}

# Plotting functions
plot_county_time_series <- function(time_data, plot_title) {
  ggplot(time_data, aes(x = ds, y = mean_percent_change, color = county_name)) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1) +
    scale_y_continuous(
      labels = scales::label_number(suffix = "%", accuracy = 1)
    ) +
    labs(
      title = plot_title,
      x = NULL,
      y = "% Change vs Baseline",
      color = "County"
    ) +
    theme_minimal()
}

## Facetted Option
plot_county_time_series_faceted <- function(time_data, plot_title) {
  ggplot(time_data, aes(x = ds, y = mean_percent_change, group = county_name)) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1) +
    facet_wrap(~ county_name, scales = "free_y") +
    scale_y_continuous(
      labels = scales::label_number(suffix = "%", accuracy = 1)
    ) +
    labs(
      title = plot_title,
      x = NULL,
      y = "% Change vs Baseline"
    ) +
    theme_minimal() + 
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

}

```


```{r fig-time-series1, fig.cap = "There are strong difference in population changes across different counties", output = T}
high_time <- make_county_time_data(high_counties)

plot_county_time_series(
  high_time,
  "% Population Density Change Over Time: Highest-Change Counties"
)


average_time <- make_county_time_data(average_counties)

plot_county_time_series(
  average_time,
  "% Population Density Change Over Time: Average-Change Counties"
)


low_time <- make_county_time_data(low_counties)

plot_county_time_series(
  low_time,
  "% Population Density Change Over Time: Lowest-Change Counties"
)

```
Check why lowest tracks so closely on Feb 9th and investigate whether outliers are skewing mean
```{r fig-time-series, fig.cap = "There are strong difference in population changes across different counties", output = T}
high_time <- make_county_time_data(high_counties)

plot_county_time_series_faceted(
  high_time,
  "% Population Density Change Over Time: Highest-Change Counties"
)


average_time <- make_county_time_data(average_counties)

plot_county_time_series_faceted(
  average_time,
  "% Population Density Change Over Time: Average-Change Counties"
)


low_time <- make_county_time_data(low_counties)

plot_county_time_series_faceted(
  low_time,
  "% Population Density Change Over Time: Lowest-Change Counties"
)

```

```{r}
#-----------Investigation of mean v median---------------
# Might not even use but want thoughts on which to use, not many outliers present but there will be a lot of 0s pulling mean down
# Facetted plotting function using median percent change
plot_county_time_series_faceted_median <- function(time_data, plot_title) {
  ggplot(time_data, aes(x = ds, y = median_percent_change, group = county_name)) +
    geom_hline(yintercept = 0, linewidth = 0.3) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1) +
    facet_wrap(~ county_name, scales = "free_y") +
    scale_y_continuous(
      labels = scales::label_number(suffix = "%", accuracy = 1)
    ) +
    labs(
      title = plot_title,
      subtitle = "County-level median percent change from baseline",
      x = NULL,
      y = "Median % Change vs Baseline"
    ) +
    theme_minimal()
}

high_time <- make_county_time_data(high_counties)

plot_county_time_series_faceted_median(
  high_time,
  "% Population Density Change Over Time: Highest-Change Counties"
)


average_time <- make_county_time_data(average_counties)

plot_county_time_series_faceted_median(
  average_time,
  "% Population Density Change Over Time: Average-Change Counties"
)


low_time <- make_county_time_data(low_counties)

plot_county_time_series_faceted_median(
  low_time,
  "% Population Density Change Over Time: Lowest-Change Counties"
)
```

## Places



```{r tbl-summary-table, tbl.cap = "Bottom 20 county-to-county crisis movement flows", output = T}
tk_mp_data_admin <- read_csv(here("Outputs", "tk_movement_between_places_aggregated_admin.csv"),
                             show_col_types = F)
county_flows <- tk_mp_data_admin |>
  group_by(start_county_name, end_county_name) |>
  summarise(
    total_crisis_movement = sum(n_crisis, na.rm = TRUE),
    avg_z_score = mean(z_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(total_crisis_movement))

county_flows |>
  slice_head(n = 20) |>
  gt()

```

# Demographic Profile

```{r}

make_demographics_table <- function(file_name,
                                    map_date = NULL,
                                    date_col = "ds",
                                    value_col = "percent_change",
                                    n_counties = 10,
                                    output_table_name = NULL,
                                    table_title = "Demographic Profile of Most Impacted Counties", desc = T,
                                    source) {
  
  # Read file from Outputs folder
  df <- read_csv(
    here("Outputs", file_name),
    show_col_types = FALSE
  )
  
  # Convert date column
  df <- df |>
    mutate("{date_col}" := as.Date(.data[[date_col]]))
  
  # Use latest date by default
  if (is.null(map_date)) {
    map_date <- max(df[[date_col]], na.rm = TRUE)
  } else {
    map_date <- as.Date(map_date)
  }
  
  # Summarize to county level
  county_demo <- df |>
    filter(.data[[date_col]] == map_date) |>
    group_by(
      county_geoid,
      county_name,
      county_state
    ) |>
    summarise(
      mean_percent_change = mean(.data[[value_col]], na.rm = TRUE),
      total_population = first(na.omit(total_population)),
      median_income = first(na.omit(median_income)),
      poverty_rate = first(na.omit(poverty_rate)),
      pct_age_65_plus = first(na.omit(pct_age_65_plus)),
      pct_no_vehicle = first(na.omit(pct_no_vehicle)),
      .groups = "drop"
    ) |>
    mutate(
      abs_change = abs(mean_percent_change)
    )
  
  # Create gt table
  demo_table <- county_demo |>
    select(
      County = county_name,
      State = county_state,
      `Population Density Change` = mean_percent_change,
      `Total Population` = total_population,
      `Median Household Income` = median_income,
      `Poverty Rate` = poverty_rate,
      `% Age 65+` = pct_age_65_plus,
      `% No Vehicle Access` = pct_no_vehicle
    )
  demo_table <- arrange(demo_table, `Population Density Change`)
  if (desc){
    demo_table <- arrange(demo_table, desc(`Population Density Change`))
  }
  demo_table |>
  head(n_counties) |>
  gt() |>
  tab_header(
    title = table_title,
    subtitle = paste0("Source: ", source)
  ) |>
  # Wrap columns
  cols_width(
    County ~ px(120),
    State ~ px(120),
    `Population Density Change` ~ px(70),
    `Total Population` ~ px(70),
    `Median Household Income` ~ px(70),
    everything() ~ px(50)
  ) |>
  tab_options(
    table.font.size = px(9),
    heading.title.font.size = px(16),
    heading.subtitle.font.size = px(9),
    data_row.padding = px(25),
    column_labels.padding = px(0),
    table.width = pct(100)
  ) |>
  fmt_number(
    columns = `Population Density Change`,
    decimals = 1,
    suffixing = FALSE
  ) |>
  fmt_number(
    columns = `Total Population`,
    decimals = 0,
    use_seps = TRUE
  ) |>
  fmt_currency(
    columns = `Median Household Income`,
    currency = "USD",
    decimals = 0
  ) |>
  fmt_percent(
    columns = c(`Poverty Rate`, `% Age 65+`, `% No Vehicle Access`),
    decimals = 1
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  opt_table_font(
    font = list(
      google_font("Roboto Condensed"),
      default_fonts()
    )
  )
}
```

```{r, output = T}
make_demographics_table(
  file_name = "tk_facebook_pop_aggregated_bing.csv",
  map_date = "2026-02-02",
  n_counties = 15,
  source = "Bing Tiles",
  table_title = "Counties with Most Increased Population Density on 2026-02-02"
) 
```

```{r, output = T}
make_demographics_table(
  file_name = "tk_facebook_pop_aggregated_bing.csv",
  n_counties = 15,
  map_date = "2026-02-02",
  desc = F,
  source = "Bing Tiles",
  table_title = "=Counties with Most Decreased Population Density on 2026-02-02"
) 
```

```{r, output = T}
make_demographics_table(
  file_name = "tk_facebook_pop_aggregated_bing.csv",
  map_date = "2026-02-12",
  n_counties = 15,
  source = "Bing Tiles",
  table_title = "Counties with Most Increased Population Density on 2026-02-12"
) 
```

```{r, output = T}
make_demographics_table(
  file_name = "tk_facebook_pop_aggregated_bing.csv",
  n_counties = 15,
  map_date = "2026-02-12",
  desc = F,
  source = "Bing Tiles",
  table_title = "Counties with Most Decreased Population Density on 2026-02-12"
) 
```

